---
title: "dplyr exercise"
author: "Shiva Raghuvanshi"
date: "12/15/2022"
output: html_document
---
# Explore Data Manipulation in R with Dplyr package

Load dplyr package,
load flights dataset,
view the hflights dataset

```{r setup, include=FALSE}
# load library package
library(dplyr)
library(hflights)
```



```{r pressure, echo=FALSE}
data(hflights)
head(hflights)
```
 
 Create a local data frame with tbl_df()
```{r}
flights <- tbl_df(hflights)
head(flights)
```
 
# Dplyr Filter: Works on rows 
Explore filtering data with dplyr and compare it with Base R commands
 
```{r}
#Base R commands for filtering all the rows for January 1
flights[flights$Month == 1 & flights$DayofMonth == 1, ]

# Dplyr commands for filtering January 1 data
filter(flights, Month == 1, DayofMonth == 1 )
```

```{r}
# Use dplyr to filter flights from Month January with Destination "DFW"
filter(flights, Month == 1, Dest == "DFW") 

# Use %in% operator for filtering data as well
filter(flights, UniqueCarrier %in% c("AA", "UA"))

# Show the last few rows of data after filtering 
tail(filter(flights, Month %in% c("1", "2"), Dest %in% c("DFW", "MIA")))

```

# select: Filter the specified columns

```{r}
# Lets see Base R code for selecting DepTime, ArrTime, and FlightNum column
flights[, c('DepTime', 'ArrTime', 'FlightNum')]

# Now, Dplyr way to select columns
select(flights, DepTime, ArrTime, FlightNum)
```

# Piping with %>%:  Perform multiple operations
```{r}
# Look at the Flight Carriers that has Departure Delay of > 60 min
filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 60)

# Now use piping
flights %>%
  select(UniqueCarrier, DepDelay) %>%
  filter(DepDelay > 60) %>%
  arrange(DepDelay) # Arrange in increasing order
```

```{r}
# Calculate Euclidean Distance between two Vectors

y <- 1:10
x <- 3:30
(x-y)^2
sum((x-y)^2)
sqrt(sum((x-y)^2)) 

#Dplyr
(x-y)^2 %>%
  sum() %>%
  sqrt()

```

# arrange:
```{r}
# Base R commands for ascending or descending ordr of data
flights[order(flights$DepDelay), c("UniqueCarrier", "DepDelay")]

#Dplyr
select(flights, UniqueCarrier, DepDelay) %>%
  arrange(DepDelay)

# Descending
select(flights, UniqueCarrier, DepDelay) %>%
  arrange(desc(DepDelay))
```

# mutate: Add new variables
```{r}
# Base R to add new variable/column "speed" in mph
flights$speed <- flights$Distance / flights$AirTime*60
flights[, c("Distance", "AirTime", "speed")]

#mutate
flights %>%
  select(Distance, AirTime) %>%
  mutate(speed = Distance/AirTime*60)

# Now store the new variable "speed" into flights data
flights <- flights %>%
  mutate(speed = Distance/AirTime*60)
```

# summarise: its narrows down the groups into values,such as mean,std dev,average...

```{r}
head(flights)
```



```{r}
#base R
head(with(flights, tapply(ArrDelay, Dest, mean, na.rm = TRUE)))

```

```{r}
# summarise

flights %>%
  group_by(Dest) %>%
  summarise(avg_delay = mean(ArrDelay, na.rm = TRUE))
  
```

```{r}
# For UniqueCarrier, find mean arrival and departure delays using summarise

flights %>%
  select(UniqueCarrier, ArrDelay, DepDelay) %>%
  group_by(UniqueCarrier) %>%
  summarise(avg_arr_delay = mean(ArrDelay, na.rm = TRUE), avg_dep_delay = mean(DepDelay, na.rm = TRUE))

```
```{r}
# For UniqueCarrier, find min and max arrival and departure delays using summarise

flights %>%
  group_by(UniqueCarrier) %>%
  summarise_each(funs(min(., na.rm=TRUE), max(., na.rm = TRUE)), ends_with('Delay'))
```

# Count
```{r}
# for each day of the Year, count total number of flights, sort in descending order
 flights %>%
   group_by(DayofMonth, Month) %>%
   count(DayofMonth, Month, sort = TRUE)
 
 # OR
 flights %>%
   group_by(DayofMonth, Month) %>%
   summarise(total_flights = n()) %>%
   arrange(desc(total_flights))

```
some more summarize exercise
```{r}
# for each destination, count total flights and distinct planes
flights
flights %>%
  group_by(Dest) %>%
  summarise(total_dest_flights = n(), plane_count = n_distinct(TailNum))
```

```{r}
# For each destination, show the number of canceled and not canceled flights
flights %>%
  group_by(Dest) %>%
  select(Cancelled) %>%
  table() %>%
  head()


```

# Windows Functions
```{r}
#For each carrir, calculate 2 days with longest Departure Delay

flights %>%
  group_by(UniqueCarrier) %>%
  select(Month, DayofMonth, DepDelay) %>%
  top_n(2) %>%
  arrange(UniqueCarrier,desc(DepDelay))
  
```

# offset functions (lead and lag)

```{r}
# for each month, calculate number of flights and and change from previous month
flights %>% 
  group_by(Month) %>%
  summarise(flight_count = n()) %>%
  mutate(change = flight_count - lag(flight_count))
```
# Random Sample
```{r}
# randomly sample the 5 rows without replacement
flights %>% sample_n(5, )
#randomly sample a fraction of data with replacement
flights %>% sample_frac(0.2, replace = TRUE)
```

# Str vs Glimpse
```{r}
glimpse(flights)
```



